---
title: "NGS: Gene-set Analysis"
author: "Omid & Guy Matz (guyeric.matz)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_width: 8
    fig_height: 8
    extra_dependencies: ["amsmath", "pifont"]
  html_document:
    df_print: paged
toc: true
toc-depth: 5
geometry: margin=1in, top=0.75in, bottom=0.75in
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Biological Question
"Compare control and treated patients at the 6th & 24h time point.
The information is stored in the `$sra.sample_attributes` in colData() object."

## Define Required Packages

```{r Packages}
req_pkgs <- c(
    # grammar of data manipulation
    # https://dplyr.tidyverse.org/
    "dplyr",

    # Tools for Splitting, Applying and Combining Data
    # https://cran.r-project.org/web/packages/plyr/index.html
    "plyr",

    # Exploratory Data Analysis and Normalization for RNA-Seq
    # https://bioconductor.org/packages/devel/bioc/vignettes/EDASeq/inst/doc/EDASeq.html
    "EDASeq",

    # provides a visual exploratory tool on correlation matrix
    # https://www.rdocumentation.org/packages/corrplot/versions/0.95
    "corrplot",

    # Unified plotting tools for statistics commonly used, such as GLM, time
    # series, PCA families, clustering and survival analysis
    # https://cran.r-project.org/web/packages/ggfortify/index.html
    "ggfortify",

    # https://bioconductor.org/packages/release/bioc/html/edgeR.html
    # Differential expression analysis of sequence count data
    "edgeR",

    # https://www.bioconductor.org/packages/release/bioc/html/genefilter.html
    # methods for filtering genes from high-throughput experiments
    "genefilter",
    
    # https://bioconductor.org/packages/release/bioc/html/limma.html
    # Linear Models for Microarray and Omics Data
    "limma",

    # https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html
    # A universal enrichment tool for interpreting omics data
    "clusterProfiler",

    # https://bioconductor.org/packages/release/bioc/html/pathview.html
    # a tool set for pathway based data integration and visualization
    "pathview",

    # https://bioconductor.org/packages/release/bioc/html/AnnotationDbi.html
    # Manipulation of SQLite-based annotations in Bioconductor
    "AnnotationDbi",

    # https://bioconductor.org/packages/release/data/annotation/html/org.Hs.eg.db.html
    # Genome wide annotation for Human, primarily based on mapping using Entrez Gene identifiers
    "org.Hs.eg.db",

    # https://bioconductor.org/packages/release/bioc/html/ReactomePA.html
    # Reactome Pathway Analysis
    "ReactomePA",

    # https://bioconductor.org/packages/release/bioc/html/enrichplot.html
    # Visualization of Functional Enrichment Result
    "enrichplot",

    # https://cran.r-project.org/web/packages/ggnewscale
    # Multiple Fill and Colour Scales in 'ggplot2'
    "ggnewscale",
    
    # https://cran.r-project.org/web/packages/ggupset
    # Combination Matrix Axis for 'ggplot2' to Create 'UpSet' Plots
    "ggupset",

    # https://cran.r-project.org/web/packages/ggridges
    # Ridgeline Plots in 'ggplot2'
    "ggridges",

    # https://bioconductor.org/packages/release/bioc/html/enrichplot.html
    # Visualization of Functional Enrichment Result
    "enrichplot",

    # https://cran.r-project.org/web/packages/knitr
    # A General-Purpose Package for Dynamic Report Generation in R
    "knitr",

    # https://bioconductor.org/packages/release/bioc/html/DESeq2.html
    # Differential gene expression analysis based on the negative binomial distribution
    "DESeq2",

    # https://cran.r-project.org/web/packages/pheatmap
    # Pretty Heatmaps
    "pheatmap",

    # https://cran.r-project.org/web/packages/cluster
    # "Finding Groups in Data": Cluster Analysis Extended Rousseeuw et
    "cluster"
)
print("Done!")
```

## Install Required Packages
```{r InstallPackages, warning=FALSE, message=FALSE, results='hide'}
if (!require("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
    BiocManager::install(version = "3.22")
}

for (pkg in req_pkgs) {
    if (!require(pkg, character.only=TRUE, quietly = TRUE)) {
        BiocManager::install(pkg)
    }
    #print(paste("Loading ", pkg))
    library(pkg, character.only=TRUE)
}

print("Done!")
```

## Load Data
```{r LoadData}
# Download data file from Google Drive, if necessary
# Then load the data
google_drive_file_id <- "1la5_fzzoEVEf5sLdpW8iD2S-eHr3DPfr"
file_url = paste0("https://drive.google.com/uc?export=download&id=",
                  google_drive_file_id)
local_data_file <- "SRP027258.rds"
server_data_file <- paste0("/opt/omicsdata/datasets")

if (! file.exists(local_data_file) && ! file.exists(server_data_file)) {
  download.file(file_url, local_data_file)
}

data <- readRDS(local_data_file)
print("Done!")
```

## Let's Give Our Data a Cleaning
```{r}
# This does nothing since we have data in all rows
sum(is.na(data$sra.sample_attributes))
```

### We identify genes that have all zero counts across all samples and remove them

#### Before
```{r}
print(table(rowSums(assay(data, "raw_counts")) == 0))
```

#### After
```{r}
# We keep the genes with the sum > 0
data_pre = data
data <- data[rowSums(assay(data, "raw_counts")) > 0, ]
print(table(rowSums(assay(data, "raw_counts")) == 0))
```

## After cleaning the data, we add columns for . . .
Group - Normal vs Tumoral
Treatment - DMSO vs SFN
Timing - 6hr vs 24hr

```{r}
data$cell_line <- as.factor(ifelse(grepl('Normal', data$sra.sample_attributes),
                               "normal", "tumoral"))
data$treatment <- as.factor(ifelse(grepl('DMSO', data$sra.sample_attributes),
                                   "dmso", "sfn"))
data$harvest_time <- as.factor(ifelse(grepl('6 hrs', data$sra.sample_attributes),
                                "6", "24"))
```

We can start by visualizing the data using a boxplot. Before plotting, it is often necessary to transform the count matrix using a `log(counts + 1)` transformation. This transformation is important because:

  - RNA-seq samples often have different sequencing depths. Log transformation stabilizes the variance and makes the distributions more comparable across samples;
  - most genes have low counts, and a few have very high counts. Log transformation makes the data more balanced and easier to analyze;
  - adding 1 in log(counts + 1) allows us to transform zero counts, which would otherwise be undefined;
  - improve interpretability.
  
```{r}
raw_counts <- assay(data, "raw_counts")
# To make readable the boxplot on all we reduce the string of column names.
colnames(raw_counts) <- substr(colnames(raw_counts), 4, 10)

# For faster plotting, use only the first 20 samples
boxplot(raw_counts[, 1:36])
boxplot(log1p(raw_counts)[, 1:36])

features = c("cell_line", "treatment", "harvest_time")
for (f in features) {
  d <- data@colData[f][, 1]
  boxplot(log1p(raw_counts)[, 1:20], title = f, col = col, xlab = "Sample",
          ylab = "log1p expression", ylim = c(0, 20),  las = 2, cex.axis = 0.7)
  legend("top", legend = c(levels(d)), col = 1:length(levels(d)),
         pch = 20, cex = 0.7, ncol = 2)
}
```